### 2006-02-02 inst/paper/paper2.tex
Made a few changes to the paper.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@16065 bc3139a8-67e5-0310-9ffc-ced21a209358

\newcommand{\Rfunction}[1]{{\textbf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\textsf{#1}}

@@

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{January 25, 2006} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

First definitions of linked views, interactive views and linked data
sets are given in the following paragraphs.  Linked views mean that if a
component that represents data (such as a point on a plot or a row in a
spreadsheet) changes its appearance on one view, then the corresponding
component on a second view also changes its appearance.  For components on
different views to be corresponding, the data displayed in those components
must come from the same entity in a model.  As an example, suppose that the
model consists of the gene expression levels from four samples.  If the same
one hundred genes are studied in the four samples, then the model could be
represented as a matrix of four columns by one hundred rows with four columns
for the four samples and one hundred rows for the one hundred genes.  In this
model, the entities are the genes.  If two scatterplots are created as views
of this model, where the first plot displays sample one versus sample two and
the second plot displays sample three versus sample four, then the points on
each plot that referred to the same gene are corresponding components.  Linked
views, which have a long history in statistics, are now considered a standard
feature of interactive data visualization software \cite{GGobi}.  Linked views
have been implemented in several pieces of statistical software, such as
XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
is based on the model-view-controller (MVC) design \cite{DesignPatterns},
which is a widely used and well understood paradigm.  The MVC design is
discussed in more detail in Section \ref{Ssec:Design}.

@@

Creating linked, interactive views of data is flexible for exploring
multivariate data because many different views can be created, such as
scatterplots, spreadsheets, and heatmaps, and these views are linked because
they are based on the same underlying data.  Thus, users can decide which
views best visually represent their data.  Also, since the views are
interactive, users can change the views while looking at them to make the
views more informative about the underlying structure in the data.  This
flexibility in linked, interactive views gives users a powerful tool for
visually exploring their data.  However, linked views can not solve all
problems.  Clearly, only a few linked views can be viewed simultaneously by
the user because only a certain number of views can be shown on a computer
screen.  Thus, there is a limit to the number of dimensions that can be
represented by linked views.

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in Step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In Step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In Step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatterplot, the first step, indicated by a circled one,
      is to tell data set 2 to update.  Then all views of data set 2 must be
      updated.  Next any data that are linked to data set 2 must be notified
      that data set 2 changed.  This information from data set 2 must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set 1 is notified that it should
      update in response to the change in data set 2 and then all views of
      data set 1 are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

The \Rclass{sPlotView} class represents a scatterplot view and it has the
extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
\Rslot{dfRows} are the row names from the model that are shown in the plot,
\Rslot{xvar} is the variable name from the model that is shown as the x
variable in the plot, and \Rslot{yvar} is the variable name from the model
that is shown as the y variable in the plot.  The \Rclass{graphView} class
represents a graph plot view and it contains no extra slots besides those in
the \Rclass{plotView} class.  The \Rclass{heatmapView} represents a heatmap
view of the data and it contains the slot, \Rslot{ordering}, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the ordering in the row and column dendrograms.  Note that some of these
views are only applicable for certain model types.  For example, the
\Rclass{graphView} class only makes sense for a view of a graph object,
which is stored in the \Rclass{graphModel} class.

Since all views are expected to be interactive, all view classes have methods
that correspond to user interaction.  Interactive views mean a response
occurs when the view has focus and an event occurs.  The events we are most
interested in responding to include key press
events, where a certain configuration of keys are pressed; button click
events, where one of the mouse buttons is clicked over the view; and mouse
movement events, where the cursor moves over the view.  Recall from Section
\ref{Ssec:Backg} that when an event occurs Gtk emits a signal that is caught
by the signal handler.  The signal handler then calls the callback function
and it is in the callback function that the corresponding method is called.
For example, when the button click event occurs on a scatterplot view, the
callback function calls the scatterplot's \Rfunction{clickEvent} method.
Using the information that is passed to the \Rfunction{clickEvent} method in
the \Rslot{where} parameter, the method converts the Gtk pixel coordinates
that indicate where the event occurred into user coordinates and then into an
object from the plot, such as a point on the plot.

@@

As shown in Figure \ref{Fig:ContWin}, the menus on the control window are
File, Edit, Display, Event, and Window.  These menus always appear on the
control window, but the menu items under each menu change depending on
the type of model that is active.  Because the goal is to have multiple
linked \Robject{MVC} objects, more than one \Robject{MVC} object is loaded and
thus, more than one model. However, the control window menus and the command
line functions only perform operations on one \Robject{MVC} object at a time.
Thus, in the \Rpackage{iSNetwork} package there is the concept of the
``active'' \Robject{MVC} object, where the ``active'' \Robject{MVC} object
(and thus, the ``active'' model) means that operations performed through the
control window menus or through the command line functions affect that
\Robject{MVC} object.  For example, the Display menu allows users to create
views of the active model, but only certain views are appropriate for a type
of model, as mentioned in 
Section \ref{Ssec:OneViews}.  If the active model was a data frame model
(\Rclass{dfModel}), then the views that the user could create are a
scatterplot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
but a graph plot (\Rclass{graphView}) does not make sense.  Thus, when the
model in the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the
Display 
menu has menu items to create a scatterplot and a spreadsheet.  Also, if
the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the command
line functions only let the user create a scatterplot and a spreadsheet view
of the active model.

@@

An example of message passing within a \Robject{MVC} object is shown in Figure
\ref{Fig:MPwithin}.  In this example, the response to a user performing a
button click event over a point in the scatterplot is to color the point red.
Setting the response to an event was discussed in Section
\ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
scatterplot, which is shown as Step 1 (the circled 1 on the figure).  In the
figure, the point that has been clicked has an arrow next to it.  Then, in
Step 2, an update data message is sent to the model, which updates
the \Rslot{virtualData} slot of the model object.  After the
model has been updated, the views must be notified that a
change has occurred so in Step 3, an update view message is created and sent
to all views of this model.  In response to this
message, the view updates itself in Step 4 and now the
point is colored red on the view to reflect the change in the
model.  In this example, the \Rslot{from} slot in the
\Robject{gUpdateDataMessage} object is the name of this \Robject{MVC} object
since the message came from within the \Robject{MVC}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatterplot.  Then an update data message is sent to the model
      in the second step, which alters the model.  Next in the third step all
      views of this model are sent an update view message, which causes the
      view to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{newMP.jpg}}
    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
      interacts with the y versus x scatterplot, the first step, indicated by
      a circled one, is to send an update data message to model 2.  After
      model 2 has been updated, all views of this model must be notified, which
      is shown in the second step when an update view message is sent.  Next
      any data that are linked to model 2 must be notified that model 2
      changed, which is shown in the third step as a send parent message.
      This message is converted into useful information for model 1 in
      the fourth step by the \Rfunction{toParent} function, and this
      information is sent to model 1 in the fifth step when an update data
      message is sent.  Finally, after model 1 is updated, an update view
      message is sent in the sixth step to update all views of model 1. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

Figure \ref{Fig:MessPass} is very similar to Figure \ref{Fig:firstMP} shown in
Section \ref{Ssec:Backg}.  Now, however, Figure \ref{Fig:MessPass} shows how
\Robject{MVC} objects are defined and how messages are passed within one
\Robject{MVC} as well as how this information gets passed on to its parent
\Robject{MVC} using messages.  Suppose for this example that when the user
clicks a mouse button over the scatterplot, the response to this event is to
color the point under the cursor 
red.  In this picture, the message passing starts when a user clicks a point,
that is indicated by the arrow, on the scatterplot.  An update data message is
sent to Model 2 in response to this user interaction.  As soon as Model 2 is
updated, it sends an update view message to all its views so that the views
are updated to reflect the changed data.  Now the scatterplot point that was
initially clicked is red.  Then a send parent message is sent because MVC
2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
update data message into an update data message that the parent \Robject{MVC}
understands.  Then the update data message that was created by the
\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
been updated, an update view message is sent to all of Model 1's views.  It is
in this last step that the view of Model 1 has a point colored red to indicate
which data from Model 1 is linked to data from Model 2.  All of the data and
the views are synchronized.

>>>>

\newcommand{\Rfunction}[1]{{\textsf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\texttt{#1}}

@@

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{February 2, 2006} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

First definitions of linked views, interactive views and linked data
sets are given in the following paragraphs.  Linked views mean that if a
component that represents data (such as a point on a plot or a row in a
spreadsheet) changes its appearance on one view, then the corresponding
component on a second view also changes its appearance.  For components on
different views to be corresponding, the data displayed in those components
must come from the same entity in a model.  As an example, suppose that the
model consists of the gene expression levels from four samples.  If the same
one hundred genes are studied in the four samples, then the model could be
represented as a matrix of four columns by one hundred rows with four columns
for the four samples and one hundred rows for the one hundred genes.  In this
model, the entities are the genes.  If two scatter plots are created as views
of this model, where the first plot displays sample one versus sample two and
the second plot displays sample three versus sample four, then the points on
each plot that referred to the same gene are corresponding components.  Linked
views, which have a long history in statistics, are now considered a standard
feature of interactive data visualization software \cite{GGobi}.  Linked views
have been implemented in several pieces of statistical software, such as
XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
is based on the model-view-controller (MVC) design \cite{DesignPatterns},
which is a widely used and well understood paradigm.  The MVC design is
discussed in more detail in Section \ref{Ssec:Design}.

@@

Creating linked, interactive views of data is flexible for exploring
multivariate data because many different views can be created, such as
scatter plots, spreadsheets, and heatmaps, and these views are linked because
they are based on the same underlying data.  Thus, users can decide which
views best visually represent their data.  Also, since the views are
interactive, users can change the views while looking at them to make the
views more informative about the underlying structure in the data.  This
flexibility in linked, interactive views gives users a powerful tool for
visually exploring their data.  However, linked views can not solve all
problems.  Clearly, only a few linked views can be viewed simultaneously by
the user because only a certain number of views can be shown on a computer
screen.  Thus, there is a limit to the number of dimensions that can be
represented by linked views.

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatter plot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
scatter plot view that is should update itself in response to the change in the
data.  Next, in Step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In Step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In Step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set 2 to update.  Then all views of data set 2 must be
      updated.  Next any data that are linked to data set 2 must be notified
      that data set 2 changed.  This information from data set 2 must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set 1 is notified that it should
      update in response to the change in data set 2 and then all views of
      data set 1 are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

The \Rclass{sPlotView} class represents a scatter plot view and it has the
extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
\Rslot{dfRows} are the row names from the model that are shown in the plot,
\Rslot{xvar} is the variable name from the model that is shown as the x
variable in the plot, and \Rslot{yvar} is the variable name from the model
that is shown as the y variable in the plot.  The \Rclass{graphView} class
represents a graph plot view and it contains no extra slots besides those in
the \Rclass{plotView} class.  The \Rclass{heatmapView} represents a heatmap
view of the data and it contains the slot, \Rslot{ordering}, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the ordering in the row and column dendrograms.  Note that some of these
views are only applicable for certain model types.  For example, the
\Rclass{graphView} class only makes sense for a view of a graph object,
which is stored in the \Rclass{graphModel} class.

Since all views are expected to be interactive, all view classes have methods
that correspond to user interaction.  Interactive views mean a response
occurs when the view has focus and an event occurs.  The events we are most
interested in responding to include key press
events, where a certain configuration of keys are pressed; button click
events, where one of the mouse buttons is clicked over the view; and mouse
movement events, where the cursor moves over the view.  Recall from Section
\ref{Ssec:Backg} that when an event occurs Gtk emits a signal that is caught
by the signal handler.  The signal handler then calls the callback function
and it is in the callback function that the corresponding method is called.
For example, when the button click event occurs on a scatter plot view, the
callback function calls the scatter plot's \Rfunction{clickEvent} method.
Using the information that is passed to the \Rfunction{clickEvent} method in
the \Rslot{where} parameter, the method converts the Gtk pixel coordinates
that indicate where the event occurred into user coordinates and then into an
object from the plot, such as a point on the plot.

@@

As shown in Figure \ref{Fig:ContWin}, the menus on the control window are
File, Edit, Display, Event, and Window.  These menus always appear on the
control window, but the menu items under each menu change depending on
the type of model that is active.  Because the goal is to have multiple
linked \Robject{MVC} objects, more than one \Robject{MVC} object is loaded and
thus, more than one model. However, the control window menus and the command
line functions only perform operations on one \Robject{MVC} object at a time.
Thus, in the \Rpackage{iSNetwork} package there is the concept of the
``active'' \Robject{MVC} object, where the ``active'' \Robject{MVC} object
(and thus, the ``active'' model) means that operations performed through the
control window menus or through the command line functions affect that
\Robject{MVC} object.  For example, the Display menu allows users to create
views of the active model, but only certain views are appropriate for a type
of model, as mentioned in 
Section \ref{Ssec:OneViews}.  If the active model was a data frame model
(\Rclass{dfModel}), then the views that the user could create are a
scatter plot (\Rclass{sPlotView}) and a spreadsheet (\Rclass{spreadView}),
but a graph plot (\Rclass{graphView}) does not make sense.  Thus, when the
model in the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the
Display 
menu has menu items to create a scatter plot and a spreadsheet.  Also, if
the ``active'' \Robject{MVC} is of class \Rclass{dfModel}, then the command
line functions only let the user create a scatter plot and a spreadsheet view
of the active model.

@@

An example of message passing within a \Robject{MVC} object is shown in Figure
\ref{Fig:MPwithin}.  In this example, the response to a user performing a
button click event over a point in the scatter plot is to color the point red.
Setting the response to an event was discussed in Section
\ref{Ssec:OneEvent}.  Here the user first clicks on a point in the
scatter plot, which is shown as Step 1 (the circled 1 on the figure).  In the
figure, the point that has been clicked has an arrow next to it.  Then, in
Step 2, an update data message is sent to the model, which updates
the \Rslot{virtualData} slot of the model object.  After the
model has been updated, the views must be notified that a
change has occurred so in Step 3, an update view message is created and sent
to all views of this model.  In response to this
message, the view updates itself in Step 4 and now the
point is colored red on the view to reflect the change in the
model.  In this example, the \Rslot{from} slot in the
\Robject{gUpdateDataMessage} object is the name of this \Robject{MVC} object
since the message came from within the \Robject{MVC}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{newMPwithin.jpg}}
    \caption{ Message Passing within a \Robject{MVC} object.  The first step,
      indicated by a circled one, occurs when a user interacts with the y
      versus x scatter plot.  Then an update data message is sent to the model
      in the second step, which alters the model.  Next in the third step all
      views of this model are sent an update view message, which causes the
      view to be updated in the fourth step. }
    \label{Fig:MPwithin}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{newMP.jpg}}
    \caption{ Message Passing Between and Within \Robject{MVCs}.  When a user
      interacts with the y versus x scatter plot, the first step, indicated by
      a circled one, is to send an update data message to model 2.  After
      model 2 has been updated, all views of this model must be notified, which
      is shown in the second step when an update view message is sent.  Next
      any data that are linked to model 2 must be notified that model 2
      changed, which is shown in the third step as a send parent message.
      This message is converted into useful information for model 1 in
      the fourth step by the \Rfunction{toParent} function, and this
      information is sent to model 1 in the fifth step when an update data
      message is sent.  Finally, after model 1 is updated, an update view
      message is sent in the sixth step to update all views of model 1. }
    \label{Fig:MessPass}
  \end{center}
\end{figure}

Figure \ref{Fig:MessPass} is very similar to Figure \ref{Fig:firstMP} shown in
Section \ref{Ssec:Backg}.  Now, however, Figure \ref{Fig:MessPass} shows how
\Robject{MVC} objects are defined and how messages are passed within one
\Robject{MVC} as well as how this information gets passed on to its parent
\Robject{MVC} using messages.  Suppose for this example that when the user
clicks a mouse button over the scatter plot, the response to this event is to
color the point under the cursor 
red.  In this picture, the message passing starts when a user clicks a point,
that is indicated by the arrow, on the scatter plot.  An update data message is
sent to Model 2 in response to this user interaction.  As soon as Model 2 is
updated, it sends an update view message to all its views so that the views
are updated to reflect the changed data.  Now the scatter plot point that was
initially clicked is red.  Then a send parent message is sent because MVC
2, which is the \Robject{MVC} object for Model 2, has a parent \Robject{MVC},
which is MVC 1.  Next the \Rfunction{toParent} function converts the child's
update data message into an update data message that the parent \Robject{MVC}
understands.  Then the update data message that was created by the
\Rfunction{toParent} function is sent to Model 1.  Finally, after Model 1 has
been updated, an update view message is sent to all of Model 1's views.  It is
in this last step that the view of Model 1 has a point colored red to indicate
which data from Model 1 is linked to data from Model 2.  All of the data and
the views are synchronized.
